---
title: "REFERENCE GENES IN *Heliconius erato lativitta*"
author: "Angel Ojeda"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_height: 4
    fig_width: 6
    highlight: tango
    theme: cerulean
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: yes
subtitle: "DATA ANALYSIS"
---

```{r setup, include=FALSE}
rm(list = ls())
options(scipen = 999, encoding = "utf-8")

pkgs <- c("knitr", "rmarkdown", "readxl", "dplyr", "ggplot2", "ggpubr",
          "rstatix", "tidyr", "car", "broom", "purrr", "scales", "kableExtra",
          "patchwork")

invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
  library(p, character.only = TRUE)
}))

opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
opts_knit$set(width = 75)
options(max.print = 200)
options(knitr.table.format = "html")   # ensures cell_spec HTML renders correctly
```

```{r helper-functions, include=FALSE}
# ── Significance codes (vectorized) ──────────────────────────────────────────
sig_code <- function(p) {
  dplyr::case_when(
    is.na(p)  ~ "NA",
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    p < 0.1   ~ ".",
    TRUE      ~ "ns"
  )
}

# ── Freedman–Diaconis optimal binwidth ────────────────────────────────────────
fd_binwidth <- function(x) {
  x <- x[!is.na(x)]
  2 * IQR(x) / length(x)^(1 / 3)
}

# ── Mean ± SD summary by groups ──────────────────────────────────────────────
data_summary <- function(data, varname, groupnames) {
  data %>%
    group_by(across(all_of(groupnames))) %>%
    summarise(
      !!paste0(varname, "_mean") := mean(.data[[varname]], na.rm = TRUE),
      !!paste0(varname, "_sd")   := sd(.data[[varname]],   na.rm = TRUE),
      .groups = "drop"
    )
}

# ── kableExtra formatted table with significance coloring ────────────────────
sig_table <- function(df, caption = NULL) {
  sig_idx  <- which(names(df) %in% c("Signif", "signif"))[1]
  sig_vals <- df[[sig_idx]]

  bg <- dplyr::case_when(
    sig_vals == "***" ~ "#c0392b",
    sig_vals == "**"  ~ "#e67e22",
    sig_vals == "*"   ~ "#f39c12",
    sig_vals == "."   ~ "#95a5a6",
    TRUE              ~ "#ecf0f1"
  )
  txt <- ifelse(sig_vals %in% c("***", "**", "*"), "white", "black")

  df %>%
    kable(caption = caption, align = "c", digits = 4,
          format.args = list(scientific = TRUE)) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, font_size = 12) %>%
    column_spec(sig_idx, color = txt, background = bg) %>%
    scroll_box(width = "100%")
}

# ── Distribution plots (density + histogram) ─────────────────────────────────
#   Returns a ggarrange composite figure
plot_distribution <- function(data, x_var, group_var,
                               title_prefix, x_label,
                               fill_color = "#FF6666", ncols = 2) {
  groups <- unique(data[[group_var]])
  plots  <- lapply(groups, function(grp) {
    df <- data %>% filter(.data[[group_var]] == grp)
    bw <- fd_binwidth(df[[x_var]])
    ggplot(df, aes(x = .data[[x_var]])) +
      geom_density(color = "black", linewidth = 0.6,
                   alpha = 0.25, fill = fill_color) +
      geom_histogram(aes(y = after_stat(density)), binwidth = bw,
                     colour = "black", fill = "white", alpha = 0.3) +
      labs(title = paste(title_prefix, grp),
           x = x_label, y = "Density") +
      scale_y_continuous(labels = scientific_format()) +
      theme_classic() +
      theme(axis.line  = element_line(color = "black", linewidth = 0.7),
            plot.title = element_text(hjust = 0.5, size = 11))
  })
  do.call(ggarrange, c(plots,
                        list(ncol = ncols,
                             nrow = ceiling(length(plots) / ncols),
                             common.legend = TRUE, legend = "bottom")))
}

# ── Levene test loop over groups ─────────────────────────────────────────────
run_levene <- function(data, response_var, group_var, formula_rhs) {
  groups  <- unique(data[[group_var]])
  results <- lapply(groups, function(grp) {
    df  <- data %>% filter(.data[[group_var]] == grp)
    fml <- as.formula(paste(response_var, "~", formula_rhs))
    res <- leveneTest(fml, data = df)
    data.frame(
      Group   = as.character(grp),
      F_value = round(res$`F value`[1], 4),
      Pr_F    = res$`Pr(>F)`[1],
      Signif  = sig_code(res$`Pr(>F)`[1]),
      stringsAsFactors = FALSE
    )
  })
  result <- do.call(rbind, results)
  names(result)[1] <- group_var
  result
}

# ── Fit a named list of GLMs ──────────────────────────────────────────────────
fit_glm_list <- function(data, response_var, group_var,
                          formula_rhs, family) {
  groups <- unique(data[[group_var]])
  models <- lapply(groups, function(grp) {
    df  <- data %>% filter(.data[[group_var]] == grp)
    fml <- as.formula(paste(response_var, "~", formula_rhs))
    glm(fml, data = df, family = family)
  })
  names(models) <- as.character(groups)
  models
}

# ── ANOVA Type II LR test from a named list of GLMs ──────────────────────────
#   Uses LR chi-square test for both Gamma and Gaussian GLMs.
#   Note: car::Anova() applied to glm objects always returns LR Chisq columns
#   regardless of family, so F-test must be handled via lm() objects instead.
#   For gaussian(link="identity"), LR and F-test conclusions are equivalent.
glm_list_anova <- function(model_list, group_var) {
  results <- lapply(names(model_list), function(nm) {
    mod    <- model_list[[nm]]
    aov_df <- as.data.frame(
      Anova(mod, type = "II", test.statistic = "LR"))
    data.frame(
      Group    = nm,
      Term     = rownames(aov_df),
      Df       = aov_df$Df,
      LR_Chisq = round(aov_df$`LR Chisq`, 3),
      Pr_Chi   = aov_df$`Pr(>Chisq)`,
      Signif   = sig_code(aov_df$`Pr(>Chisq)`),
      stringsAsFactors = FALSE, row.names = NULL
    )
  })
  result <- do.call(rbind, results)
  names(result)[1] <- group_var
  result
}

# ── QQ-plots from a named list of GLMs ───────────────────────────────────────
plot_qq <- function(model_list, ncols = 2) {
  plots <- lapply(names(model_list), function(nm) {
    res <- residuals(model_list[[nm]])
    ggqqplot(res,
             xlab = "Theoretical Quantiles",
             ylab = "Sample Quantiles",
             ggtheme = theme_pubr()) +
      ggtitle(paste("Q-Q Residuals:", nm)) +
      theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
            axis.title = element_text(size = 10),
            axis.text  = element_text(size = 9))
  })
  do.call(ggarrange, c(plots,
                        list(ncol = ncols,
                             nrow = ceiling(length(plots) / ncols),
                             common.legend = TRUE, legend = "bottom")))
}

# ── Residual histograms from a named list of GLMs ────────────────────────────
plot_residuals <- function(model_list, ncols = 2) {
  plots <- lapply(names(model_list), function(nm) {
    res <- residuals(model_list[[nm]])
    bw  <- fd_binwidth(res)
    ggplot(data.frame(res = res), aes(x = res)) +
      geom_density(color = "black", linewidth = 0.6,
                   alpha = 0.8, fill = "lightblue") +
      geom_histogram(aes(y = after_stat(density)), binwidth = bw,
                     colour = "black", fill = "white", alpha = 0.3) +
      labs(title = paste("Residuals:", nm),
           x = "Residuals", y = "Density") +
      theme_classic() +
      theme(axis.line  = element_line(color = "black", linewidth = 0.7),
            plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
            axis.title = element_text(size = 10),
            axis.text  = element_text(size = 9))
  })
  do.call(ggarrange, c(plots,
                        list(ncol = ncols,
                             nrow = ceiling(length(plots) / ncols),
                             common.legend = TRUE, legend = "bottom")))
}

# ── AIC comparison: additive vs interaction ───────────────────────────────────
compare_aic <- function(data, response_var, group_var,
                         formula_add, formula_int, family) {
  groups  <- unique(data[[group_var]])
  results <- lapply(groups, function(grp) {
    df      <- data %>% filter(.data[[group_var]] == grp)
    mod_add <- glm(as.formula(paste(response_var, "~", formula_add)),
                   data = df, family = family)
    mod_int <- glm(as.formula(paste(response_var, "~", formula_int)),
                   data = df, family = family)
    aic_a   <- round(AIC(mod_add), 2)
    aic_i   <- round(AIC(mod_int), 2)
    data.frame(
      Group           = as.character(grp),
      AIC_Additive    = aic_a,
      AIC_Interaction = aic_i,
      Delta_AIC       = round(aic_a - aic_i, 2),
      Preferred       = ifelse(aic_i < aic_a - 2, "Interaction", "Additive"),
      stringsAsFactors = FALSE
    )
  })
  result <- do.call(rbind, results)
  names(result)[1] <- group_var
  result
}

# ── Auto family candidates based on data range ────────────────────────────────
#   Gamma requires strictly positive Y. If Y has zeros or negatives, only
#   Gaussian(identity) is returned — no AIC selection is needed.
auto_family_candidates <- function(y) {
  if (any(y <= 0, na.rm = TRUE)) {
    list("Gaussian(identity)" = gaussian(link = "identity"))
  } else {
    list(
      "Gamma(log)"         = Gamma(link = "log"),
      "Gamma(inverse)"     = Gamma(link = "inverse"),
      "Gaussian(identity)" = gaussian(link = "identity")
    )
  }
}

# ── AIC-based family selection per group ──────────────────────────────────────
#   Fits all valid candidate families for each group using the additive formula.
#   Convergence warnings are suppressed (informative but not critical for AIC).
#   If only one family is valid, returns NULL (no selection needed).
family_selection_aic <- function(data, response_var, group_var, formula_rhs) {
  candidates <- auto_family_candidates(data[[response_var]])

  if (length(candidates) == 1L) return(NULL)   # auto-selected, skip

  groups  <- unique(data[[group_var]])
  results <- lapply(groups, function(grp) {
    df <- data %>% filter(.data[[group_var]] == grp)

    aics <- sapply(names(candidates), function(fam_name) {
      mod <- tryCatch(
        suppressWarnings(
          glm(as.formula(paste(response_var, "~", formula_rhs)),
              data = df, family = candidates[[fam_name]])
        ),
        error = function(e) NULL
      )
      if (is.null(mod)) Inf else AIC(mod)
    })

    row              <- as.data.frame(t(round(aics, 2)))
    row[[group_var]] <- as.character(grp)
    row$Best         <- names(candidates)[which.min(aics)]
    row
  })

  result <- do.call(rbind, results)
  result[, c(group_var, names(candidates), "Best")]
}

# ── Consensus family: family that wins in the most groups ─────────────────────
get_consensus_family <- function(aic_table, data, response_var) {
  candidates  <- auto_family_candidates(data[[response_var]])
  best_counts <- sort(table(aic_table$Best), decreasing = TRUE)
  cname       <- names(best_counts)[1]
  list(
    name    = cname,
    family  = candidates[[cname]],
    counts  = best_counts,
    n_total = nrow(aic_table)
  )
}

# ── Format AIC table for kable: highlight winner per row in green ─────────────
format_family_table <- function(aic_df, group_var) {
  fam_cols <- setdiff(names(aic_df), c(group_var, "Best"))
  display  <- aic_df

  # Convert AIC columns to character to accept cell_spec HTML
  display[, fam_cols] <- lapply(display[, fam_cols], as.character)

  for (i in seq_len(nrow(aic_df))) {
    row_vals <- as.numeric(aic_df[i, fam_cols])
    min_col  <- fam_cols[which.min(row_vals)]

    for (col in fam_cols) {
      val <- round(as.numeric(aic_df[i, col]), 2)
      if (!is.finite(val)) {
        display[i, col] <- cell_spec("—", color = "#95a5a6", italic = TRUE)
      } else if (col == min_col) {
        display[i, col] <- cell_spec(val, background = "#27ae60",
                                      color = "white", bold = TRUE)
      } else {
        display[i, col] <- as.character(val)
      }
    }
  }
  display
}

# ── Barplot function for EGR (worst / best genes) ────────────────────────────
#   quality_label is shown as a styled badge inside the plot (top-left),
#   so it integrates with the design without replacing any existing element.
barplot_egr <- function(data_egr, quality_label,
                         y_label = "Relative Gene Expression — log(EGR)") {
  df <- data_summary(data_egr, varname = "Log_EGR",
                     groupnames = c("Tissue", "Treatment", "Sex")) %>%
    mutate(Tissue = dplyr::recode(Tissue,
                                  "A" = "Abdomen",
                                  "H" = "Head",
                                  "T" = "Thorax"))

  y_min    <- min(df$Log_EGR_mean - df$Log_EGR_sd, na.rm = TRUE)
  y_max    <- max(df$Log_EGR_mean + df$Log_EGR_sd, na.rm = TRUE)
  y_limits <- c(floor(y_min), ceiling(y_max))

  ggplot(df, aes(x = Tissue, y = Log_EGR_mean, fill = Treatment)) +
    geom_bar(stat = "identity", color = "black",
             position = position_dodge()) +
    geom_errorbar(aes(
      ymin = ifelse(Log_EGR_mean >= 0, Log_EGR_mean, Log_EGR_mean - Log_EGR_sd),
      ymax = ifelse(Log_EGR_mean >= 0, Log_EGR_mean + Log_EGR_sd, Log_EGR_mean)),
                  width = 0.2, position = position_dodge(0.9)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    # ── Quality badge: dark teal label in top-left of plot area ──────────────
    annotate("label",
             x          = -Inf,
             y          = Inf,
             label      = toupper(quality_label),
             hjust      = -0.15,
             vjust      = 1.4,
             fontface   = "bold",
             size       = 4.5,
             fill       = "#264653",
             color      = "white",
             label.size = 0,
             label.padding = unit(0.35, "lines")) +
    facet_grid(. ~ Sex) +
    scale_fill_manual(values = c("#264653", "#b5b5b5", "#f4a261")) +
    scale_y_continuous(limits = y_limits,
                       breaks = seq(y_limits[1], y_limits[2], 1)) +
    labs(fill = "Treatment",
         y    = y_label,
         x    = "Tissue") +
    theme(
      panel.background = element_blank(),
      panel.border     = element_rect(color = "black", fill = NA),
      strip.background = element_rect(fill = "gray90", color = "black"),
      strip.text       = element_text(size = 14, face = "bold"),
      axis.title       = element_text(size = 14),
      axis.text        = element_text(size = 12),
      legend.title     = element_text(size = 12),
      legend.text      = element_text(size = 10)
    )
}
```

---

Angel Ojeda  - `angel.ojeda@ikiam.edu.ec` - Universidad Regional Amazónica IKIAM, Tena (Ecuador)

Paola Calderón  - `pcalderon@stri.pa` - Instituto Smithsoniano de investigaciones tropicales, Gamboa (Panamá)

Joshué Ruiz  - `joshue.ruiz@ikiam.edu.ec` - Universidad Regional Amazónica IKIAM, Tena (Ecuador)

Leo Camino  - `lcamino@stri.pa` - Instituto Smithsoniano de investigaciones tropicales, Gamboa (Panamá)

Shane Wright  - `dswright@bio.lmu.de` - Munich University (Germany)

Caroline Bacquet - `caroline.bacquet@ikiam.edu.ec` - Universidad Regional Amazónica IKIAM, Tena (Ecuador)

---

```{r load-data, include=FALSE}
# NOTE: This script lives in analysis/. Data lives in ../data/All_data.xlsx
# (relative to this file) — knitr sets the working directory to the Rmd's
# own location automatically, so no setwd() is needed.

dir.create("../outputs", showWarnings = FALSE)   # Save all figures here (gitignored)

# ── Ct data ─────────────────────────────────────────────────────────────────
data_ct <- read_excel("../data/All_data.xlsx", sheet = "Cq_tagged") %>%
  mutate(
    Gene      = as.factor(Gene),
    Treatment = as.factor(Treatment),
    Sex       = as.factor(Sex),
    Tissue    = as.factor(Tissue),
    Ct        = as.numeric(Ct)
  ) %>%
  na.omit()

str(data_ct)

# ── EGR data ─────────────────────────────────────────────────────────────────
data_EGR <- read_excel("../data/All_data.xlsx", sheet = "EGR") %>%
  mutate(
    Gene      = as.factor(Gene),
    Quality   = as.factor(Quality),
    Treatment = as.factor(Treatment),
    Sex       = as.factor(Sex),
    Tissue    = as.factor(Tissue),
    EGR       = as.numeric(EGR),
    Log_EGR   = as.numeric(Log_EGR)
  ) %>%
  na.omit()

str(data_EGR)

# ── Figure height variables (computed once, reused in chunk options) ──────────
n_genes       <- length(unique(data_ct$Gene))
fig_h_ct      <- ceiling(n_genes / 2) * 3.5   # ~3.5 in per row of 2 panels
n_quality     <- length(unique(data_EGR$Quality))
fig_h_egr_bar <- n_quality * 5.0              # ~5 in per quality group
```

# 1. Ct Analysis

## 1.1. Ct Summary

```{r ct-summary, echo=FALSE}
summary(data_ct) %>%
  kable(caption = "Summary statistics — Ct values") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  scroll_box(width = "100%")
```

## 1.2. Ct Boxplot

####### I made changes in this part, because we agreed to split the information in three boxplots i.e sex, tissue and treatment####

```{r ct-boxplot, fig.width=10, fig.height=6}
p_box <- ggplot(data_ct, aes(x = Gene, y = Ct,  fill = Treatment)) +
  geom_boxplot(outlier.shape = NA, color = "black") +
  labs(x = "Gene", y = "Ct") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(face= 2, size = 13, color = "black"),
    axis.title.y = element_text(size =16, face = 2),
    axis.title.x = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = "black"),
    legend.title = element_text(face = "bold", size =12),
    legend.position = "right",
    legend.text = element_text(size= 10),
    axis.ticks.x = element_blank(),
    axis.ticks = element_line(colour = "black", linewidth = 1)
  ) +
  scale_fill_manual(values = c("Cold" = "#118ab2", "Heat" = "#ff9f1c", "Control" = "#8d99ae"))

################################################################################

condition_order <- c("H", "T", "A")
data_ct$Tissue <- factor(data_ct$Tissue, levels = condition_order)
boxplot_tissue<- ggplot(data_ct, aes(x = Gene, y = Ct, fill = Tissue)) +
   geom_boxplot(outlier.shape = NA, colour = "black", linewidth = 0.5) + 
  
  labs(
    title = " ",
    x = "Genes",
    y = "Ct"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(face= 2, size = 13, color = "black"),
    axis.title.y = element_text(size =16, face = 2),
    axis.title.x = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = "black"),
    legend.title = element_text(face = "bold", size =12),
    legend.position = "right",
    legend.text = element_text(size= 10),
    axis.ticks.x = element_blank(),
    axis.ticks = element_line(colour = "black", linewidth = 1)
  ) +
  scale_fill_manual(values = c("H" = "#457b9d", "A" = "#f26419", "T" = "#f4d35e" ))


print(boxplot_tissue)
###########################################################
boxplot_sex<- ggplot(data_ct, aes(x = Gene, y = Ct, fill = Sex)) +
   geom_boxplot(outlier.shape = NA, colour = "black", linewidth = 0.5) +  
  labs(
    title = " ",
    x = "Genes",
    y = "Ct"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(face="italic", size = 13, color = "black"),
    axis.text.y = element_text(face= 2, size = 13, color = "black"),
    axis.title.y = element_text(size =15, face = 2),
    axis.title.x = element_text(size =15, face = 2),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = "black"),
    legend.title = element_text(face = "bold", size =12),
    legend.position = "right",
    legend.text = element_text(size= 10),
    axis.ticks = element_line(colour = "black", linewidth = 1)
  ) +
  scale_fill_manual(values = c("Male" = "#606c38", "Female" = "#bc6c25"))


print(boxplot_sex)
###############################################################################
combined_plot <- p_box / boxplot_tissue / boxplot_sex +
  plot_layout(heights = c(1, 1, 1)) 
print(combined_plot)


ggsave("../outputs/01_Ct_boxplot.png", plot = combined_plot, width = 10, height = 6, dpi = 400)
```


## 1.3. Ct Distribution

```{r ct-distribution, fig.width=10, fig.height=fig_h_ct}
p_ct_dist <- plot_distribution(
  data         = data_ct,
  x_var        = "Ct",
  group_var    = "Gene",
  title_prefix = "Ct distribution —",
  x_label      = "Ct"
)

print(p_ct_dist)
ggsave("../outputs/02_Ct_distribution.png", plot = p_ct_dist,
       width = 10, height = fig_h_ct, dpi = 400)
```

### 1.3.1. Ct Homogeneity of Variance (Levene Test)

> Tests whether Ct variance is homogeneous across Treatment × Sex × Tissue for each gene.
> A **significant result** (p < 0.05) indicates heteroscedasticity — relevant for model selection.

```{r ct-levene}
ct_levene <- run_levene(
  data         = data_ct,
  response_var = "Ct",
  group_var    = "Gene",
  formula_rhs  = "Treatment * Sex * Tissue"
)

sig_table(ct_levene, caption = "Levene Test — Ct homogeneity of variance by gene")
```

## 1.4. Ct GLM Analysis

> **Model specification:** `Ct ~ predictors` with family selected automatically
> by AIC comparison in section 1.4.0 (consensus across all genes).
> Type II Likelihood Ratio (LR) tests are used for all terms.

### 1.4.0. GLM Family Selection

> The optimal GLM family for Ct is selected **per gene** by comparing AIC across
> three candidate specifications fitted with the additive formula.
> The **consensus family** (winner in most genes) is then used automatically in
> all hypothesis-testing models (sections 1.4.1–1.4.3).
> This table is intended for **Supplementary Material**.
>
> Candidate families: `Gamma(log)`, `Gamma(inverse)`, `Gaussian(identity)`.
> All are valid because Ct values are strictly positive (range: 
> `r round(min(data_ct$Ct), 1)` – `r round(max(data_ct$Ct), 1)`).
> The winning AIC per gene is highlighted in **green**.

```{r ct-family-selection-fit, include=FALSE}
ct_family_aic <- family_selection_aic(
  data         = data_ct,
  response_var = "Ct",
  group_var    = "Gene",
  formula_rhs  = "Treatment + Sex + Tissue"
)

consensus_ct <- get_consensus_family(ct_family_aic, data_ct, "Ct")
```

```{r ct-family-selection-table}
format_family_table(ct_family_aic, "Gene") %>%
  kable(format   = "html",
        escape   = FALSE,
        caption  = "Supplementary — Family selection AIC per gene (additive model)",
        align    = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

> **Consensus family selected: `r consensus_ct$name`**  
> (`r max(consensus_ct$counts)` of `r consensus_ct$n_total` genes).
> All models in sections 1.4.1–1.4.3 use this family automatically.

### 1.4.1. Interaction Model — `Treatment * Sex * Tissue`

```{r ct-glm-interaction-fit, include=FALSE}
models_ct_int <- fit_glm_list(
  data         = data_ct,
  response_var = "Ct",
  group_var    = "Gene",
  formula_rhs  = "Treatment * Sex * Tissue",
  family       = consensus_ct$family
)
```


#### ANOVA Results (Type II LR Test)

```{r ct-anova-interaction}
ct_anova_int <- glm_list_anova(models_ct_int, group_var = "Gene")
sig_table(ct_anova_int, caption = "Interaction GLM — ANOVA Type II LR test (Ct ~ Treatment * Sex * Tissue)")
```

#### Heatmap — P-values by Gene and Condition

> P-values extracted automatically from the interaction GLM results above.
> Colour scale: **red** = low p-value (significant), **blue** = high p-value (non-significant).
> Significance markers: `***` p < 0.001 · `**` p < 0.01 · `*` p < 0.05.
> Labels use scientific notation when p < 0.001, otherwise 3 decimal places.
> Genes are ordered by their maximum p-value across conditions (most stable at top).
> Palette: RdYlBu (colorblind-friendly, as recommended by Lin).

```{r ct-heatmap, fig.width=10, fig.height=6}
# ── Build heatmap data from GLM interaction results ───────────────────────────
term_labels <- c(
  "Treatment"              = "Treatment",
  "Sex"                    = "Sex",
  "Tissue"                 = "Tissue",
  "Treatment:Sex"          = "Treat*Sex",
  "Treatment:Tissue"       = "Treat*Tissue",
  "Sex:Tissue"             = "Sex*Tissue",
  "Treatment:Sex:Tissue"   = "Treat*Sex*Tissue"
)

condition_order <- c("Treatment", "Sex", "Tissue",
                     "Treat*Sex", "Treat*Tissue",
                     "Sex*Tissue", "Treat*Sex*Tissue")

heatmap_data <- ct_anova_int %>%
  filter(Term %in% names(term_labels)) %>%
  mutate(
    Condition = factor(term_labels[Term], levels = condition_order),
    Gene      = as.character(Gene)
  ) %>%
  select(Gene, Condition, Pr_Chi)

# Order genes by maximum p-value descending
gene_order <- heatmap_data %>%
  group_by(Gene) %>%
  summarise(max_p = max(Pr_Chi, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(max_p)) %>%
  pull(Gene)

heatmap_data$Gene <- factor(heatmap_data$Gene, levels = gene_order)

# ── Labels and significance annotation ───────────────────────────────────────
heatmap_data <- heatmap_data %>%
  mutate(
    # P-value label: scientific if < 0.001, else 3 decimal places
    p_label = ifelse(
      Pr_Chi < 0.001,
      formatC(Pr_Chi, format = "e", digits = 2),
      formatC(round(Pr_Chi, 3), format = "f", digits = 3)
    ),
    # Significance stars on a second line below the p-value
    sig_label = dplyr::case_when(
      Pr_Chi < 0.001 ~ "***",
      Pr_Chi < 0.01  ~ "**",
      Pr_Chi < 0.05  ~ "*",
      TRUE           ~ ""
    ),
    # Combined label: p-value + stars (newline separator)
    full_label = ifelse(
      sig_label == "",
      p_label,
      paste0(p_label, "\n", sig_label)
    )
  )

# ── Plot ──────────────────────────────────────────────────────────────────────
p_heatmap <- ggplot(heatmap_data,
                    aes(x = Condition, y = Gene, fill = Pr_Chi)) +
  # Base tiles — uniform border, no significance overlay
  geom_tile(colour = "black", linewidth = 0.4) +
  # Combined p-value + stars label
  geom_text(aes(label = full_label),
            size = 3.2, color = "black", lineheight = 0.85) +
  scale_fill_distiller(
    palette   = "RdYlBu",
    direction = 1,          # low p (significant) → red; high p → blue
    limits    = c(0, 1),
    name      = "P-value"
  ) +
  scale_x_discrete(limits = condition_order) +
  labs(x = "Conditions", y = "Gene") +
  theme_minimal() +
  theme(
    axis.text.y       = element_text(color = "black", size = 11, face = "italic"),
    axis.text.x       = element_text(size = 11, angle = 45,
                                      vjust = 1, hjust = 1, color = "black"),
    axis.title.y      = element_text(size = 14, face = "bold"),
    axis.title.x      = element_text(size = 14, face = "bold"),
    axis.ticks        = element_line(colour = "black", linewidth = 0.8),
    axis.ticks.length = unit(6, "pt"),
    legend.text       = element_text(size = 10),
    legend.title      = element_text(face = "bold", size = 14),
    panel.background  = element_rect(fill = "white", color = "black"),
    panel.grid        = element_blank()
  )

print(p_heatmap)
ggsave("../outputs/03_Ct_heatmap_interaction.png", plot = p_heatmap,
       width = 10, height = 6, dpi = 400)
```



#### Q-Q Plots of Residuals

```{r ct-qq-interaction, fig.width=10, fig.height=fig_h_ct}
p_qq_int <- plot_qq(models_ct_int, ncols = 2)
print(p_qq_int)
ggsave("../outputs/05_Ct_QQ_interaction.png", plot = p_qq_int,
       width = 10, height = fig_h_ct, dpi = 400)
```

#### Residual Histograms

```{r ct-resid-interaction, fig.width=10, fig.height=fig_h_ct}
p_res_int <- plot_residuals(models_ct_int, ncols = 2)
print(p_res_int)
ggsave("../outputs/06_Ct_residuals_interaction.png", plot = p_res_int,
       width = 10, height = fig_h_ct, dpi = 400)
```

### 1.4.2. Model Comparison — AIC (Additive vs. Interaction)

> **Interpretation:** ΔAIC > 2 favours the interaction model; |ΔAIC| < 2 indicates
> equivalent support and the simpler (additive) model is preferred by parsimony.
> This table is provided as supplementary justification for model selection.

```{r ct-aic}
ct_aic <- compare_aic(
  data         = data_ct,
  response_var = "Ct",
  group_var    = "Gene",
  formula_add  = "Treatment + Sex + Tissue",
  formula_int  = "Treatment * Sex * Tissue",
  family       = consensus_ct$family
)

ct_aic %>%
  kable(caption = "AIC comparison — Additive vs Interaction Ct GLMs",
        align = "c", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  column_spec(which(names(ct_aic) == "Preferred"),
              bold = TRUE,
              color = ifelse(ct_aic$Preferred == "Interaction",
                             "#c0392b", "#27ae60")) %>%
  scroll_box(width = "100%")
```

```{r ct-aic-interpretation, results='asis'}
# ── Automatic AIC interpretation paragraph ────────────────────────────────────
n_add  <- sum(ct_aic$Preferred == "Additive")
n_int  <- sum(ct_aic$Preferred == "Interaction")
n_tot  <- nrow(ct_aic)

# Genes where interaction wins
int_genes <- ct_aic$Gene[ct_aic$Preferred == "Interaction"]
int_genes_txt <- if (length(int_genes) == 0) {
  "none"
} else {
  paste(paste0("*", int_genes, "*"), collapse = ", ")
}

# Overall conclusion
if (n_add >= n_tot * 0.75) {
  conclusion <- paste0(
    "The additive model was preferred in **", n_add, " of ", n_tot,
    "** genes (|ΔAIC| > 2), indicating that interactions between ",
    "Treatment, Sex, and Tissue do not explain substantial additional ",
    "variance in Ct values. This supports interpreting the effects of ",
    "each factor independently."
  )
} else if (n_int >= n_tot * 0.75) {
  conclusion <- paste0(
    "The interaction model was preferred in **", n_int, " of ", n_tot,
    "** genes (|ΔAIC| > 2), suggesting that the effect of at least one ",
    "factor on Ct depends on the levels of another factor."
  )
} else {
  conclusion <- paste0(
    "Model preference was mixed: the additive model was favoured in **",
    n_add, "** genes and the interaction model in **", n_int,
    "** of ", n_tot, " genes. ",
    if (length(int_genes) > 0)
      paste0("Genes showing interaction support: ", int_genes_txt, ".")
    else ""
  )
}

cat("> 📊 **Automatic AIC summary:** ", conclusion, "\n")
```


---

# 2. Relative Expression Analysis (EGR)

> **Context:** EGR values result from RefFinder / geNorm rankings. The `Quality`
> variable classifies genes as "worst" or "best" reference genes based on their
> stability scores. The goal is to compare how expression results change when
> normalizing with each set.

## 2.1. EGR Summary

```{r egr-summary}
summary(data_EGR) %>%
  kable(caption = "Summary statistics — EGR data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  scroll_box(width = "100%")
```

## 2.2. EGR Distribution

### 2.2.1. Raw EGR Distribution

```{r egr-distribution, fig.width=10, fig.height=5}
p_egr_dist <- plot_distribution(
  data         = data_EGR,
  x_var        = "EGR",
  group_var    = "Quality",
  title_prefix = "EGR distribution —",
  x_label      = "EGR",
  fill_color   = "#FF6666",
  ncols        = 2
)

print(p_egr_dist)
ggsave("../outputs/07_EGR_distribution.png", plot = p_egr_dist,
       width = 10, height = 5, dpi = 400)
```

### 2.2.2. Log(EGR) Distribution

```{r egr-log-distribution, fig.width=10, fig.height=5}
p_loegr_dist <- plot_distribution(
  data         = data_EGR,
  x_var        = "Log_EGR",
  group_var    = "Quality",
  title_prefix = "log(EGR) distribution —",
  x_label      = "log(EGR)",
  fill_color   = "#66b2ff",
  ncols        = 2
)

print(p_loegr_dist)
ggsave("../outputs/08_LogEGR_distribution.png", plot = p_loegr_dist,
       width = 10, height = 5, dpi = 400)
```

## 2.3. EGR Homogeneity of Variance (Levene Test)

### 2.3.1. Raw EGR

```{r egr-levene}
egr_levene <- run_levene(
  data         = data_EGR,
  response_var = "EGR",
  group_var    = "Quality",
  formula_rhs  = "Treatment * Sex * Tissue"
)

sig_table(egr_levene,
          caption = "Levene Test — EGR homogeneity of variance by gene quality")
```

### 2.3.2. Log(EGR)

```{r egr-log-levene}
log_egr_levene <- run_levene(
  data         = data_EGR,
  response_var = "Log_EGR",
  group_var    = "Quality",
  formula_rhs  = "Treatment * Sex * Tissue"
)

sig_table(log_egr_levene,
          caption = "Levene Test — log(EGR) homogeneity of variance by gene quality")
```

## 2.4. Barplots — Worst vs Best Reference Genes

```{r egr-barplots, fig.width=10, fig.height=fig_h_egr_bar}
qualities  <- unique(data_EGR$Quality)
barplots   <- lapply(qualities, function(q) {
  barplot_egr(data_EGR %>% filter(Quality == q), quality_label = q)
})
names(barplots) <- as.character(qualities)

p_bars <- do.call(ggarrange,
                  c(barplots,
                    list(ncol = 1, nrow = length(qualities),
                         common.legend = TRUE, legend = "bottom")))

print(p_bars)
ggsave("../outputs/09_EGR_barplots.png", plot = p_bars,
       width = 10, height = fig_h_egr_bar, dpi = 400)
```

## 2.5. Log(EGR) GLM Analysis

> **Family auto-selected: `gaussian(link = "identity")`.**
> Log(EGR) contains negative values 
> (range: `r round(min(data_EGR$Log_EGR), 2)` – `r round(max(data_EGR$Log_EGR), 2)`),
> which are incompatible with Gamma distributions (requiring Y > 0 strictly).
> No AIC family comparison is needed — Gaussian is the only valid candidate.
> This is equivalent to a standard linear model; Type II LR tests are used for effects.

### 2.5.1. Interaction Model — `Treatment * Sex * Tissue`

```{r egr-glm-interaction-fit, include=FALSE}
models_egr_int <- fit_glm_list(
  data         = data_EGR,
  response_var = "Log_EGR",
  group_var    = "Quality",
  formula_rhs  = "Treatment * Sex * Tissue",
  family       = gaussian(link = "identity")
)
```

#### ANOVA Results (Type II F-Test)

```{r egr-anova-interaction}
egr_anova_int <- glm_list_anova(models_egr_int, group_var = "Quality")
sig_table(egr_anova_int,
          caption = "Interaction GLM — ANOVA Type II LR test (log(EGR) ~ Treatment * Sex * Tissue)")
```

#### Q-Q Plots of Residuals

```{r egr-qq-interaction, fig.width=10, fig.height=5}
p_egr_qq_int <- plot_qq(models_egr_int, ncols = 2)
print(p_egr_qq_int)
ggsave("../outputs/12_EGR_QQ_interaction.png", plot = p_egr_qq_int,
       width = 10, height = 5, dpi = 400)
```

#### Residual Histograms

```{r egr-resid-interaction, fig.width=10, fig.height=5}
p_egr_res_int <- plot_residuals(models_egr_int, ncols = 2)
print(p_egr_res_int)
ggsave("../outputs/13_EGR_residuals_interaction.png", plot = p_egr_res_int,
       width = 10, height = 5, dpi = 400)
```

### 2.5.2. Model Comparison — AIC (Additive vs. Interaction)

```{r egr-aic}
egr_aic <- compare_aic(
  data         = data_EGR,
  response_var = "Log_EGR",
  group_var    = "Quality",
  formula_add  = "Treatment + Sex + Tissue",
  formula_int  = "Treatment * Sex * Tissue",
  family       = gaussian(link = "identity")
)

egr_aic %>%
  kable(caption = "AIC comparison — Additive vs Interaction log(EGR) GLMs",
        align = "c", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  column_spec(which(names(egr_aic) == "Preferred"),
              bold = TRUE,
              color = ifelse(egr_aic$Preferred == "Interaction",
                             "#c0392b", "#27ae60")) %>%
  scroll_box(width = "100%")
```

```{r egr-aic-interpretation, results='asis'}
# ── Automatic AIC interpretation paragraph ────────────────────────────────────
n_add_e <- sum(egr_aic$Preferred == "Additive")
n_int_e <- sum(egr_aic$Preferred == "Interaction")
n_tot_e <- nrow(egr_aic)

int_qual <- egr_aic$Quality[egr_aic$Preferred == "Interaction"]
int_qual_txt <- if (length(int_qual) == 0) "none" else
  paste0("*", int_qual, "*", collapse = ", ")

if (n_add_e == n_tot_e) {
  egr_conclusion <- paste0(
    "The additive model was preferred for **all ", n_tot_e,
    "** gene quality groups (|ΔAIC| > 2), indicating that interactions ",
    "between Treatment, Sex, and Tissue do not improve the explanation ",
    "of log(EGR) variance. Effects of each factor can be interpreted independently."
  )
} else if (n_int_e == n_tot_e) {
  egr_conclusion <- paste0(
    "The interaction model was preferred for **all ", n_tot_e,
    "** gene quality groups, suggesting that the combined effect of ",
    "Treatment, Sex, and Tissue on log(EGR) is not additive."
  )
} else {
  egr_conclusion <- paste0(
    "Model preference differed by gene quality group: additive preferred ",
    "in **", n_add_e, "** and interaction in **", n_int_e, "** of ",
    n_tot_e, " groups. Groups favouring the interaction model: ",
    int_qual_txt, "."
  )
}

cat("> 📊 **Automatic AIC summary:** ", egr_conclusion, "\n")
```


---

```{r session-info, include=TRUE, echo=FALSE}
sessionInfo() %>%
  capture.output() %>%
  paste(collapse = "\n") %>%
  cat()
```
